---
title: "Tutorial 2: Bootstrap and Permutation Tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Bootstrap and Permutation Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ndplane)
```

## Introduction

This tutorial covers how to perform bootstrap and permutation tests to assess the uncertainty and significance of Niche Divergence Plane results.

## Setup Data

```{r data}
set.seed(456)
n <- 200
env1 <- runif(n, 0, 100)
env2 <- runif(n, 0, 100)
data_env <- data.frame(env1 = env1, env2 = env2)

# Species A
p_a <- ifelse(env1 < 50, 0.7, 0.2)
occ_a <- rbinom(n, 1, p_a)
data_a <- cbind(p = occ_a, data_env)

# Species B (similar to A, for testing null)
p_b <- ifelse(env1 < 55, 0.7, 0.2)
occ_b <- rbinom(n, 1, p_b)
data_b <- cbind(p = occ_b, data_env)
```

## Bootstrap Cloud

The bootstrap test helps visualize the uncertainty in our index estimates.

```{r bootstrap, message=FALSE, warning=FALSE}
form <- ~ env1 + env2
variables <- c("env1", "env2")

boot_res <- ndp_bootstrap(data_a, data_b, form, variables, n_iter = 20)

# Plotting the cloud for env1
boot_env1 <- boot_res[boot_res$variable == "env1", ]
plot_ndp(boot_env1, main = "Bootstrap Cloud (env1)")
```

## Permutation Test

The permutation test evaluates the null hypothesis that there is no difference between the two species' niches.

```{r permutation, message=FALSE, warning=FALSE}
perm_res <- ndp_permutation(data_a, data_b, form, variables, n_iter = 20)

# Observe the distribution of null results
perm_env1 <- perm_res[perm_res$variable == "env1", ]
plot_ndp(perm_env1, main = "Permutation Null Distribution (env1)")
```
