---
title: "Tutorial 1: Basic Niche Divergence Plane Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: Basic Niche Divergence Plane Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ndplane)
```

## Introduction

This tutorial demonstrates the basic workflow for calculating the Niche Divergence Plane (NDP) indices using Maxent-like models fitted via `glmnet_mx`.

## Data Preparation

We will use some synthetic data for two species across two environmental variables.

```{r data}
set.seed(123)
n <- 200
env1 <- runif(n, 0, 100)
env2 <- runif(n, 0, 100)
data_env <- data.frame(env1 = env1, env2 = env2)

# Species A prefers low env1
p_a <- ifelse(env1 < 40, 0.8, 0.1)
occ_a <- rbinom(n, 1, p_a)
data_a <- cbind(p = occ_a, data_env)

# Species B prefers high env1
p_b <- ifelse(env1 > 60, 0.8, 0.1)
occ_b <- rbinom(n, 1, p_b)
data_b <- cbind(p = occ_b, data_env)
```

## Modeling with `glmnet_mx`

We fit Maxent-like models for each species.

```{r modeling}
form <- ~ env1 + env2 + I(env1^2) + I(env2^2)
model_a <- glmnet_mx(data_a$p, data_a[, c("env1", "env2")], form)
model_b <- glmnet_mx(data_b$p, data_b[, c("env1", "env2")], form)
```

## Extracting Response Curves

```{r response_curves}
rc_a_env1 <- get_response_curve(model_a, "env1")
rc_b_env1 <- get_response_curve(model_b, "env1")

plot_response_curves(rc_a_env1, rc_b_env1, main = "Response to env1", xlab = "env1")
```

## Calculating NDP Indices

```{r ndp}
ov <- beta_overlap(rc_a_env1, rc_b_env1)
diss <- niche_diss(rc_a_env1, rc_b_env1, ov)
excl <- niche_excl(rc_a_env1, rc_b_env1)

indices <- data.frame(variable = "env1", dissimilarity = diss, exclusivity = excl)
print(indices)

plot_ndp(indices, labels = indices$variable)
```
